---
title: "Class 19: Pertussis Mini Project"
author: "Devanshi"
format: pdf
toc: true
date: 2022-11-30
---

# Investigating Pertussis cases by year

First install `datapasta` to be able to read the file as a dataframe. File is from https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

```{r}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L),
  Cases = c(107473,
                               164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)
```

Plotting a graph of number of cases vs years

> Q1:

```{r}
library(ggplot2)
ggplot(cdc) + aes(x = Year, y = Cases) + geom_point() + 
  geom_line() + labs(title = "Pertussis Cases by Year") 
```

## Vaccinations introduced

Add lines at vaccine year introductions (1946 and 1996) to the plot.

> Q2:

```{r}
ggplot(cdc) + aes(x = Year, y = Cases) + geom_point() + 
  geom_line() + labs(title = "Pertussis Cases by Year") + geom_vline(xintercept = 1946, linetype = "dashed", 
                                                                     color = "red") + geom_vline(xintercept = 1996, 
                                                                                                 linetype = "dashed", color = "blue")
```

> Q3:

Some reasons for this spike could be: 1. mutations in the strain 2. lesser people taking vaccines 3. the ap vaccine 'degrading' over the years as opposed to the wp vaccine

## Exploring CMI-PB data

First install the `jsonlite` package to be able to read JSon data from the CMI-PB dataset available online giving more information on the type of vaccine and demographics of children who received the vaccine.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = T)
```

> Q4: How many aP and wP infancy vaccinated subjects are there?

```{r}
table(subject$infancy_vac)
```

> Q5: How many male and female subjects are there?

```{r}
table(subject$biological_sex)
```

> Q6: What is the breakdown of race and biological sex?

```{r}
table(subject$race, subject$biological_sex)
```

## Side note: Working with Dates

```{r}
library(lubridate)
today()
```

```{r}
today() - ymd("1999-11-20")
```

```{r}
head(subject$year_of_birth, 3)
head(subject$date_of_boost, 3)
```

```{r}
subject$age <- time_length(today() - 
                             ymd(subject$year_of_birth), "years")
```

Average age of individuals:

```{r}
mean(subject$age)
summary(subject$age)
```

> Q7: Determine (i) average age of wP individuals, (ii) average age of aP individuals and (iii) if they're significantly different.

```{r}
library(dplyr)
ap <- subject %>% filter(infancy_vac == "aP")
round(mean(time_length(ap$age)))
```

```{r}
wp <- wp <- subject %>% filter(infancy_vac == "wP")
round(mean(time_length(wp$age)))
```

> Q8: Determine age during time of boost

```{r}
diff <- ymd(subject$date_of_boost) - 
  ymd(subject$year_of_birth)
age_boost <- time_length(diff, "years")
head(age_boost)
```

> Q9: With the help of a faceted boxplot or histogram, do you think these two groups are significantly different?

Making a histogram with the summary data faceted by aP and wP:

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) 
```

We can read more data from the same CMI-PB website.

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = T)
head(specimen)
```

```{r}
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = T)
head(titer)
```

We can use the `dplyr "join"` function to join these data together with our subject data.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(specimen, subject)
dim(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(meta, titer)
dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```

There were very few visits at visit 8.

## Examining IgG1 Ab titer levels

We can filter this data set by isotype

```{r}
library(dplyr)

ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

> Q13: Make a boxplot of Ab titer levels

```{r}
ggplot(ig1) + aes(x = MFI, y = antigen, col = infancy_vac) + geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2) + 
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust=1))
```

> Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

The highest is between Fim2/3 which are fimbrial proteins that are a part of the *pertussis* structure.

We can also facet wrap the plot by infancy_vac rates

```{r}
ggplot(ig1) + aes(x = MFI, y = antigen, col = infancy_vac) + geom_boxplot() + 
  facet_wrap(vars(infancy_vac, visit), nrow=2) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```


> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like.

```{r}
filter(ig1, antigen == "Measles") %>%
  ggplot() + aes(MFI, col = infancy_vac) + geom_boxplot(show.legend = F) + facet_wrap(vars(visit)) + theme_bw()
```

```{r}
filter(ig1, antigen == "FIM2/3") %>%
  ggplot() + aes(MFI, col = infancy_vac) + 
  geom_boxplot(show.legend = F) + facet_wrap(vars(visit)) + theme_bw()
```

> Q16: What do you notice about these two antigens time courses and the FIM2/3 data in particular?

The FIM2/3 levels increase over time, a lot more than that of Measles especially during visit 5. Trends are also similar for aP and wP.

> Q17: Do you see any clear difference in aP vs. wP responses?

aP and wP responses are fairly similar across Measles and FIM2/3.


## Obtaining CMI-PB RNA seq data

We can obtain RNA seq data for the IGHG1 gene. 

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

Then we can join this to our meta dataset that contains subject and specimen.

```{r}
ssrna <- inner_join(rna, meta)
```

> Q18. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) + aes(x = visit, y = tpm, group=subject_id) + geom_point() + geom_line(alpha=0.2)
```

> Q19.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

It's at its maximum level at visit 4, after which it drops in visit 5. 

> Q20. Does this pattern in time match the trend of antibody titer data? If not, why not?

This pattern does not exactly match the trend of the ab titer data. There was a spike seen in the antigen FMI2/3 levels at visit 5 but these did not reduce as time went by. This is since antibodies are longer lived where as tpm degraded by visit 5 once the antibodies had been made after visit 4. 


Coloring and faceting a boxplot by `infancy_vac` 

```{r}
ggplot(ssrna) + aes(tpm, col = infancy_vac) + geom_boxplot() + facet_wrap(vars(visit))
```

There are no clear differences between aP and wP. Looking into a particular visit:

```{r}
ssrna %>%
  filter(visit == 4) %>%
  ggplot() + aes(tpm, col = infancy_vac) + geom_density() + geom_rug()
```

