---
title: "Covid-19 Vaccine Project"
author: "Devanshi"
format: pdf
toc: true
---

#Importing vax data

```{r}
url <- "https://data.chhs.ca.gov/dataset/ead44d40-fd63-4f9f-950a-3b0111074de8/resource/ec32eece-7474-4488-87f0-6e91cb577458/download/covid19vaccinesbyzipcode_test.csv"
vax <- read.csv(url)
head(vax)
```

> Q1. What column details the total number of people fully vaccinated? 

persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? 

zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 

```{r}
min(vax$as_of_date)
```

> Q4. What is the latest date in this dataset? 

```{r}
max(vax$as_of_date)
```

```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

13

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```


> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
(sum( is.na(vax$persons_fully_vaccinated) ) / length(vax$persons_fully_vaccinated)) * 100
```


> Q8. [Optional]: Why might this data be missing?

Perhaps some people have taken just one dose of the vaccine and therefore are not fully vaccinated.

# Working with dates

```{r}
library(lubridate)
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[length(vax$as_of_date)]
```


> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

# Working with Zip codes

Install the `zipcodeR` package. 

```{r}
library(zipcodeR)

zip_distance('90024','92122')
```

Pulling data for all zip codes in this data set

```{r}
zipdata <- reverse_zipcode(vax$zip_code_tabulation_area)
```

# Focussing on the San Diego Area

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```


> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
sd$zip_code_tabulation_area[match(max(sd$age12_plus_population), sd$age12_plus_population)]
```

Filtering data as of 2022-11-15.

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”?

```{r}
latest_sd <- filter(sd, as_of_date == "2022-11-15")
mean(latest_sd$percent_of_population_fully_vaccinated, na.rm = T)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
library(ggplot2)
ggplot(latest_sd) + aes(percent_of_population_fully_vaccinated) + geom_histogram()
```

## Focusing on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla, CA 92109", x = "Date", 
       y="Percent Vaccinated")
```


## Comparing to similar sized areas

```{r}
vax_36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")
head(vax_36)

vax_36_mean <- mean(vax_36$percent_of_population_fully_vaccinated, 
                    na.rm = T)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
ggplot(ucsd) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) + geom_hline(yintercept = vax_36_mean, linetype = "dashed", col = "red") + ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla, CA 92109", 
       x = "Date", 
       y="Percent Vaccinated")

```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?

```{r}
summary(vax_36$percent_of_population_fully_vaccinated)

```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax_36) + aes(percent_of_population_fully_vaccinated) + 
  geom_histogram()
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
Both of these zipcodes' means are lower than the calculated mean for the La Jolla zipcode. 


> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax_36_all <- filter(vax, age5_plus_population > 36144)
mean_vax_36_all <- mean(vax_36_all$percent_of_population_fully_vaccinated, 
                        na.rm = T)

ggplot(vax_36_all) + aes(x = as_of_date, y = percent_of_population_fully_vaccinated, group = zip_code_tabulation_area) + geom_line(alpha=0.2, color= "blue") +  labs(x = "Date", 
                                                                                                                                                                               y = "Percent Vaccinated", title = "Vaccination rates across California", subtitle = "Only areas with a population above 36k are shown") + geom_hline(yintercept = mean_vax_36_all, linetype= "dashed")
```

> Q21. How do you feel about traveling for Thanksgiving Break and meeting for in-person class afterwards?

Not great. We need to be more careful.